---
title: "An Analysis of Lake Michigan-Huron Water Levels, 1980-2021"
author: "Eric Chen, Yanyu Long"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_depth: 2
subtitle: STATS 531 Midterm Project (W21)
---

------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE,
	include = TRUE,
	comment='',
	fig.align = "center"
)
```


```{css, include = TRUE, echo = FALSE}
body{ /* Normal */
  font-size: 15px;
}
span.math{ /* Formulas */
  font-size: 14px;
}
pre { /* Scrollable code block */
  max-height: 250px;
  overflow-y: auto;
}
.comment { /* Comments, to be deleted later */
  margin: 30px;
  padding-left: 10px;
  color: Darkred;
  border-left-style: solid;
  border-color: #f2f2f2;
  font-weight: 600;
}
```

```{r, echo = FALSE}
opar = par()
library(tidyverse)

generate_aic_table=function(data, P, Q, D=0, ...){
	table=matrix(NA, (P+1), (Q+1))
	for(p in 0:P) {
		for(q in 0:Q) {
		  model_aic = try(
		    arima(data, order = c(p, D, q), method="ML", ...)$aic, 
		    silent = TRUE
		  )
		  table[p + 1,q + 1] = ifelse(
		    inherits(model_aic, "try-error"), NA, model_aic
		  )
		}
	}
	dimnames(table) = list(paste("AR", 0:P, sep=""), paste("MA", 0:Q, sep=""))
  table
}

format_aic_tables = function(table, ...) {
  min_idx = which(table == min(table, na.rm = TRUE), arr.ind = TRUE)
  table %>% 
    knitr::kable(
      format = "html", digits = 2, escape = TRUE, ...
    ) %>%
    kableExtra::row_spec(row = 0, align = 'c') %>%
    kableExtra::column_spec(column = 1, bold = TRUE) %>%
    kableExtra::kable_styling("striped", full_width = FALSE) %>%
    kableExtra::add_footnote(sprintf(
      "Model with the lowest AIC: %s %s, AIC = %4.2f\n",
      rownames(table)[min_idx[1, 1]], colnames(table)[min_idx[1, 2]],
      min(table, na.rm = TRUE)
    ))
}

generate_acf_qqplot = function(model) {
  par(mfrow = c(1, 2))
  acf(model$resid, na.action = na.pass, 
      main = "Autocorrelation function\nestimation")
  qqnorm(as.numeric(model$resid))
  qqline(as.numeric(model$resid))
  par(opar)
}

summary_data = function(df, caption = NULL) {
  df %>%
    summary() %>%
    knitr::kable(caption = caption) %>%
    kableExtra::column_spec(column = 1:length(df)+1, width_min = "10em") %>%
    kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

## Outline

**Use full data: 1980 - 2020**

[DELETE THIS WHEN FINISHED!]{.comment}  

- 1) [Eric] introduction and motivation

- 2) [Yanyu] plot the data, describe the data

- 3) fitting an ARMA model
  - [Eric] check for a trend
  - [Eric] check for seasonality
    - spectrogram
    - trend + noise + cycles decomposition
  - [Eric] pick appropriate model class (SARMA if there is seasonality, ARIMA if there is a trend, etc.)
  
  - [Yanyu] AIC table
  - [Yanyu] goodness of fit
    - residual plot
    - ACF of residuals
    - QQ-plot / Shapiro-Wilk test for normality
  
  - [Yanyu] investigate roots
  - [Yanyu] profile likelihood?
  - [Yanyu] chi-squared test for nested models?
  
- 4) [Eric] investigate how other lakes can act as predictor for Lake Michigan
  - motivation: is there a lag in the warmer lake (Erie) that can be used for prediction? Warmer lakes may be more susceptible to climactic changes
  - plot the data, describe the data
  - cross-spectrum, cross-correlation, coherency
  - linear regression with ARMA errors 
    - goodness of fit analysis, likelihood-ratio tests, etc.
  - conduct subset of analyses from (2) on Lake Erie data
    - fit model from (2)
    - run some diagnostics / goodness-of-fit
    
- 5) [Eric] conclusion

------------

## Introduction

The Great Lakes are a major source of water, recreation, and industry for over 30 million people, including most Michigan residents. However, in recent years, there has been a growing concern over steadily rising water levels: news sources such as the [Green Bay Press-Gazette](https://www.greenbaypressgazette.com/story/news/local/oconto-county/2020/08/04/lake-michigan-breaks-34-year-old-high-water-record/3294377001/), the [Detroit Free Press](https://www.freep.com/story/news/local/michigan/2020/07/17/great-lakes-water-levels-records-erosion-damage/5450910002/), and [The Washington Post](https://www.washingtonpost.com/weather/2019/11/08/great-lakes-water-levels-have-swung-record-lows-record-highs-heres-why/) have highlighted the resulting property damage, shoreline erosion, and economic disruption. As such, environmental scientists have been investigating the potential causes and implications of high water in the Great Lakes, tying this into the larger conversation about climate change. 

In this project, we consider the variation in water levels in Lake Michigan-Huron over the time period 1980-2021, given that the majority of residents in the Great Lakes watershed live by Lake Michigan.  

----

**NB**: Lake Michigan and Lake Huron are hydrologically [one lake](https://en.wikipedia.org/wiki/Lake_Michigan%E2%80%93Huron), since the flow of water through the Straits of Mackinac keeps their respective water levels in near-equilibrium. 

------------

## Data

The data consists of monthly lake level means from February 1980 to January 2021 and was downloaded from the Great Lakes Environmental Research Laboratory (GLERL), an arm of the National Oceanic and Atmospheric Administration (NOAA). There is no missing values in the data. 


```{r, echo=FALSE}
## reading in and formatting data
data_mich = read.csv('michigan_huron.csv', colClasses = list(Date = "Date")) %>%
  rename_with(~ str_extract(.x, "([[:alpha:]]+)")) %>% # get rid of `.` in colnames
  select(c("Date", "MSL")) 
level = data_mich$MSL
```

```{r, echo=FALSE}
data_mich %>%
  summary_data(caption = "**Table 1.** Summary statistics of the lake Michigan-Huron data")
```

The water level had two noticeable peaks in October 1986 and July 1997, hit bottom in 2013, and has been steadily rising since then. 

```{r, echo=FALSE, fig.width=8, fig.height=3.5, fig.cap="**Figure 1.** Lake Michigan-Huron Water Levels, 1980-2021"}
date_breaks = with(data_mich, seq.Date(from = min(Date), to = max(Date), by = '10 years'))
data_mich %>%
  ggplot(aes(Date, MSL)) + 
    theme_bw() +
    geom_line(color="#69b3a2", size=1) +
    ylab("Mean Lake Level (ft)") +
    scale_x_date(breaks=date_breaks, date_labels='%Y')
```

[Below is the EDA of the lake Erie dataset, move to section 3 later?]{.comment}

```{r, echo=FALSE}
data_erie = read.csv("erie.csv", colClasses = list(Date = "Date")) %>%
  rename_with(~ str_extract(.x, "([[:alpha:]]+)")) %>% # get rid of `.` in colnames
  select(c("Date", "MSL"))
```

```{r, echo=FALSE}
data_erie %>%
  summary_data(caption = "**Table x.** Summary statistics of the lake Erie data")
```

The water level of Lake Erie peaked in June 1986 and June 1997, and has presented an upward trend since 2000. 

```{r, echo=FALSE, fig.width=8, fig.height=3.5, fig.cap="**Figure x.** Lake Erie Water Levels, 1980-2021"}
data_erie %>%
  ggplot(aes(Date, MSL)) + 
    theme_bw() +
    geom_line(color="#69b3a2", size=1) +
    ylab("Mean Lake Level (ft)") +
    scale_x_date(breaks=date_breaks, date_labels='%Y')
```

Plotting the centered water levels of lake Erie and lake Michigan-Huron together, we can see that the two sets of data present a similar overall trend. 

```{r, echo=FALSE, fig.width=8, fig.height=3.5, fig.cap="**Figure x.** Centered water level of Lake Erie and Lake Michigan-Huron"}
ggplot() + 
  theme_bw() +
  geom_line(data=data_erie, 
            aes(Date, MSL-mean(MSL, na.rm=T), alpha="Erie"), 
            color="#69b3a2", size=1) +
  geom_line(data=data_mich, 
            aes(Date, MSL-mean(MSL, na.rm=T), alpha="Michigan-Huron"),
            color="#69b3a2", size=1) +
  ylab("Centered Mean Lake Level (ft)") +
  scale_x_date(breaks=date_breaks, date_labels='%Y') +
  scale_alpha_manual(name="Lake", values = c("Michigan-Huron" = 0.4, "Erie" = 1)) +
  theme(legend.position="bottom", legend.margin=margin(-10))
```

------------

## Methods and Analysis

### Detrending

### Spectral Analysis

```{r, echo=FALSE, fig.width = 6, fig.height = 5}
stl(ts(level, start = c(2010, 2), frequency=12), s.window = 12) %>% plot
```

```{r, echo=FALSE, fig.width = 6, fig.height = 5}
# transform Date to numeric type
date_num = seq(from = 2010 + 1/12, length = nrow(data_mich), by = 1/12)

trend = ts(loess(level ~ date_num, span = 0.5)$fitted,
           start = c(2010, 2), frequency = 12)
noise = ts(level - loess(level ~ date_num, span = 0.1)$fitted,
           start = c(2010, 2), frequency = 12)
cycles = level - trend - noise
plot(ts.union(level, trend, noise, cycles),
     main = "Decomposition of lake levels as trend + noise + cycles")
```

### Model Selection


### Diagnostics and Goodness-of-Fit

[residual analysis, causality, invertibility, ...]{.comment}

```{r echo=FALSE}
attach(data_mich)

generate_aic_table(MSL, P=5, Q=5, D=0) %>%
  format_aic_tables(caption = "AIC of $\\mathrm{ARMA}(p, q)$ models, $0 \\leq p, q \\leq 5$")
```

```{r, echo=FALSE, fig.width = 6, fig.height = 3}
# Model 1: ARMA
m1 = arima(level, c(5, 0, 4), method='ML')
generate_acf_qqplot(m1)
params = coef(m1)
roots = list(
  ar = polyroot(c(1, -params[grep("^ar", names(params))])),
  ma = polyroot(c(1, params[grep("^ma", names(params))]))
)
lapply(roots, abs)
```

```{r, echo=FALSE}
generate_aic_table(MSL, P=5, Q=5, D=1) %>%
  format_aic_tables(caption = "AIC of $\\mathrm{ARIMA}(p, 1, q)$ models, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE, fig.width = 6, fig.height = 3}
# Model 2: ARIMA
m2 = arima(MSL, c(3, 1, 3), method='ML')
generate_acf_qqplot(m2)
params = coef(m2)
roots = list(
  ar = polyroot(c(1, -params[grep("^ar", names(params))])),
  ma = polyroot(c(1, params[grep("^ma", names(params))]))
)
lapply(roots, abs)
```

[Similarly, we can fit SARMA models and regression models with ARMA errors ...]{.comment}

```{r, echo=FALSE, fig.width = 6, fig.height = 3}
generate_aic_table(
  level, P=5, Q=5, D=0,
  seasonal = list(order = c(1, 0, 0), period = 12)
) %>%
  format_aic_tables(caption = "AIC of $\\mathrm{SARMA}(p, q) \\times (1, 0)_{12}$ models, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE, fig.width = 6, fig.height = 3}
# Model 3: SARMA
m3 = arima(level, c(5, 0, 0), method='ML',
           seasonal = list(order = c(1, 0, 0), period = 12))
generate_acf_qqplot(m3)
```



```{r echo=FALSE}
# generate_aic_table(level, P=5, Q=5, D=0, xreg=???) %>%
#   format_aic_tables(caption = "AIC of Regression with $\\mathrm{ARMA}(p, q)$ errors, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE}
## Model 4: Regression with ARMA errors
# m4 = arima(level, c(5, 0, 5), method='ML', xreg=???)
# generate_acf_qqplot(m4)
detach(data_mich)
```


```{r, echo=FALSE, fig.width=8, fig.height=4}
## plotting out fitted values
plot(level, type = "l")
lines(level - m1$resid, lty = 2, col = "red")
lines(level - m2$resid, lty = 2, col = "green")
lines(level - m3$resid, lty = 2, col = "blue")
title(main='Fitted Values of Models 1, 2, and 3')
legend(
  "topleft",
  legend = c("Original data", "Model 1", "Model 2", "Model 3"),
  col = c("black", "red", "green", "blue"),
  lty = c(1, 2, 2, 2),
  bty = "n", horiz = FALSE, ncol = 2, cex = 1
)
```


## "Extension" (Part 3)


------------


## Conclusion


------------

## References

1. https://tidesandcurrents.noaa.gov/waterlevels.html?id=9075014&type=Tide+Data&name=Harbor%20Beach&state=MI (data!)

1. https://www.epa.gov/greatlakes/facts-and-figures-about-great-lakes

1. https://en.wikipedia.org/wiki/Lake_Michigan%E2%80%93Huron


