---
title: "An Analysis of Lake Michigan-Huron Water Levels, 2010-2021"
author: "Eric Chen, Yanyu Long"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_depth: 2
subtitle: STATS 531 Midterm Project (W21)
---

------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE,
	include = TRUE,
	comment='',
	fig.align = "center"
)
```


```{css, include = TRUE, echo = FALSE}
body{ /* Normal */
  font-size: 15px;
}
span.math{ /* Formulas */
  font-size: 14px;
}
pre { /* Scrollable code block */
  max-height: 250px;
  overflow-y: auto;
}
.comment { /* Comments, to be deleted later */
  margin: 30px;
  padding-left: 10px;
  color: Darkred;
  border-left-style: solid;
  border-color: #f2f2f2;
}
```

```{r, echo = FALSE}
opar = par()
library(tidyverse)

generate_aic_table=function(data, P, Q, D=0, ...){
	table=matrix(NA, (P+1), (Q+1))
	for(p in 0:P) {
		for(q in 0:Q) {
		  table[p + 1,q + 1] = arima(data, order = c(p, D, q), method="ML")$aic
		}
	}
	dimnames(table) = list(paste("AR", 0:P, sep=""), paste("MA", 0:Q, sep=""))
  table
}

format_aic_tables = function(table, ...) {
  min_idx = which(table == min(table), arr.ind = TRUE)
  table %>% 
    knitr::kable(
      format = "html", digits = 2, escape = TRUE, ...
    ) %>%
    kableExtra::row_spec(row = 0, align = 'c') %>%
    kableExtra::column_spec(column = 1, bold = TRUE) %>%
    kableExtra::kable_styling("striped", full_width = FALSE) %>%
    kableExtra::add_footnote(sprintf(
      "Model with the lowest AIC: %s %s, AIC = %4.2f\n",
      rownames(table)[min_idx[1, 1]], colnames(table)[min_idx[1, 2]],
      min(table)
    ))
}

generate_acf_qqplot = function(model) {
  par(mfrow = c(1, 2))
  acf(model$resid, na.action = na.pass, 
      main = "Autocorrelation function estimation")
  qqnorm(as.numeric(model$resid))
  qqline(as.numeric(model$resid))
  par(opar)
}
```

## Outline

[DELETE THIS WHEN FINISHED!]{.comment}

- seasonal analysis
  - spectrogram
  - trend + noise + cycles decomposition
- trend fitting
  - ARIMA
  - linear regression with ARMA errors? (needs exogenous variable)
- fitting an ARMA model
  - use trend and seasonality to inform model (SARIMA?)
  - AIC table
  - goodness of fit
    - residual plot
    - ACF of residuals
    - QQ-plot / Shapiro-Wilk test for normality
  - investigate roots
  - profile likelihood?
  - chi-squared test for nested models?
  
- comparing to full data (1860-2021?)

------------

## Introduction

The Great Lakes are a major source of water, recreation, and industry for over 30 million people, including most Michigan residents. However, in recent years, there has been a growing concern over steadily rising water levels: news sources such as the [Green Bay Press-Gazette](https://www.greenbaypressgazette.com/story/news/local/oconto-county/2020/08/04/lake-michigan-breaks-34-year-old-high-water-record/3294377001/), the [Detroit Free Press](https://www.freep.com/story/news/local/michigan/2020/07/17/great-lakes-water-levels-records-erosion-damage/5450910002/), and [The Washington Post](https://www.washingtonpost.com/weather/2019/11/08/great-lakes-water-levels-have-swung-record-lows-record-highs-heres-why/) have highlighted the resulting property damage, shoreline erosion, and economic disruption. As such, environmental scientists have been investigating the potential causes and implications of high water in the Great Lakes, tying this into the larger conversation about climate change. 

In this project, we consider the variation in water levels in Lake Michigan-Huron over the time period $????$-2021. [include goals and objectives here] 

----

**NB**: Lake Michigan and Lake Huron are hydrologically [one lake](https://en.wikipedia.org/wiki/Lake_Michigan%E2%80%93Huron), since the flow of water through the Straits of Mackinac keeps their respective water levels in near-equilibrium. 

------------

## Data

The data consists of monthly lake level means and was downloaded from the Great Lakes Environmental Research Laboratory (GLERL), an arm of the National Oceanic and Atmospheric Administration (NOAA). We plot the data below.


```{r echo=FALSE}
## reading in and formatting data
data = read.csv('michigan_huron_level.csv')
data$Date = as.Date(data$Date)
level <- data$MSL..ft.
```


```{r, echo=FALSE, fig.width=8, fig.height=5}
data %>%
  ggplot(aes(Date, MSL..ft.)) + 
      theme_bw() +
      geom_line(color="#69b3a2", size=1) +
      # scale_color_brewer(palette = "Set3") +
      ggtitle('Lake Michigan-Huron Water Levels, 2010-2021') +
      ylab("Mean Lake Level (ft)") +
      scale_x_date(breaks='1 year', date_labels='%Y')+
      theme(axis.text.x=element_text(angle=60, hjust=1), plot.title=
            element_text(hjust=0.5)) 
```

```{r, echo=FALSE, fig.width = 6, fig.height = 5}
# plot(ts.union(ts_price, ts_import))
```

### Spectral Analysis.

[Data cleaning, exploratory analysis, spectrum analysis ... ]{.comment}

```{r, echo=FALSE, fig.width = 6, fig.height = 5}
stl(ts(level, frequency=12), s.window = 12) %>% plot
```

```{r, echo=FALSE, fig.width = 6, fig.height = 5}
# transform Date to numeric type
date_num = seq(from = 2010, length = nrow(data), by = 1/12)

trend = ts(loess(level ~ date_num, span = 0.5)$fitted, 
           start = c(2010, 6), frequency = 12)
noise = ts(level - loess(level ~ date_num, span = 0.1)$fitted, 
            start = c(2010, 6), frequency = 12)
cycles = level - trend - noise
plot(ts.union(level, trend, noise, cycles),
     main = "Decomposition of lake levels as trend + noise + cycles")
```



------------

## Methods  

[Equations, hypotheses, notations, ...]{.comment}

------------

## Results  

[Model selection, residual analysis, causality, invertibility, ...]{.comment}

```{r echo=FALSE}
attach(data)

generate_aic_table(MSL..ft., P=5, Q=5, D=0) %>% 
  format_aic_tables(caption = "AIC of $\\mathrm{ARMA}(p, q)$ models, $0 \\leq p, q \\leq 5$")
```

```{r, echo=FALSE, fig.width = 6, fig.height = 3}
# Model 1: ARMA
m1 = arima(level, c(5, 0, 4))
generate_acf_qqplot(m1)
params = coef(m1)
roots = list(
  ar = polyroot(c(1, -params[grep("^ar", names(params))])),
  ma = polyroot(c(1, params[grep("^ma", names(params))]))
)
lapply(roots, abs)
```

```{r, echo=FALSE}
generate_aic_table(MSL..ft., P=5, Q=5, D=1) %>%
  format_aic_tables(caption = "AIC of $\\mathrm{ARIMA}(p, 1, q)$ models, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE}
# Model 2: ARIMA
m2 = arima(MSL..ft., c(3, 1, 3), method='ML')
generate_acf_qqplot(m2)
params = coef(m2)
roots = list(
  ar = polyroot(c(1, -params[grep("^ar", names(params))])),
  ma = polyroot(c(1, params[grep("^ma", names(params))]))
)
lapply(roots, abs)
```

[Similarly, we can fit SARMA models and regression models with ARMA errors ...]{.comment}

```{r, echo=FALSE, fig.width = 6, fig.height = 3}
generate_aic_table(
  level, P=5, Q=5, D=0,
  seasonal = list(order = c(1, 0, 0), period = 12)
) %>%
  format_aic_tables(caption = "AIC of $\\mathrm{SARMA}(p, q) \\times (1, 0)_{12}$ models, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE}
# Model 3: SARMA
m3 = arima(level, c(5, 0, 4), seasonal = list(order = c(1, 0, 0), period = 12))
generate_acf_qqplot(m3)
```



```{r echo=FALSE}
# generate_aic_table(level, P=5, Q=5, D=0, xreg=???) %>% 
#   format_aic_tables(caption = "AIC of Regression with $\\mathrm{ARMA}(p, q)$ errors, $0 \\leq p, q \\leq 5$")
```

```{r echo=FALSE}
## Model 4: Regression with ARMA errors
# m4 = arima(level, c(5, 0, 5), xreg=???)
# generate_acf_qqplot(m4)
detach(data)
```


```{r, echo=FALSE, fig.width=8, fig.height=5}
## plotting out fitted values
plot(level, type = "l")
lines(level - m1$resid, lty = 2, col = "red")
lines(level - m2$resid, lty = 2, col = "green")
lines(level - m3$resid, lty = 2, col = "blue")
title(main='Fitted Values of Models 1, 2, and 3')
```


------------

## Conclusions  


------------

## References  

https://tidesandcurrents.noaa.gov/waterlevels.html?id=9075014&type=Tide+Data&name=Harbor%20Beach&state=MI (data!)

https://www.epa.gov/greatlakes/facts-and-figures-about-great-lakes 

https://en.wikipedia.org/wiki/Lake_Michigan%E2%80%93Huron


